---
title: "Acacia_saligna"
author: "Lilith Diener"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Temporal assessment of the density and population structure of Acacia saligna in the Cape Flats Nature Reserve

Invasive alien plants pose a serious threat to native vegetation. Knowledge of 
changes in the density of invaders is thus crucial to evaluate the effectiveness
of eradication methods. Research on the population density and structure of 
Acacia saligna was conducted between 2019 and 2024 in a previously burnt section 
of the Cape Flats Nature Reserve, and its impact on the local plant species 
richness determined. In seven 4mx4m quadrats each, the number of A. saligna 
individuals was recorded in three areas across the reserve (Ecotone, Sand Fynbos,
Strandveld). A. saligna plants were cleared and a decline in the population's 
average density was observed.

```{r, echo=TRUE}

# Install and load packages
# install.packages("tidyr")
# options("install.lock"=FALSE)
# install.packages("ggplot2")
# options("install.lock"=FALSE)

library(tidyr)
library(dplyr)
library(ggplot2)
```

Read in the dataset

```{r, echo=TRUE}
AcaciaSaligna_Data <- read.csv("AcaciaSaligna_Data.csv")
```

Remove the columns "Area" and "Number_other_species" from the dataset

```{r, echo=TRUE}
AcaciaSaligna_Data <- read.table("AcaciaSaligna_Data.csv", header = TRUE, 
                                 sep = ",", 
		colClasses = c("NULL","integer", "character", "integer", "integer", 
						               "numeric", 
										"integer", "numeric", "integer", "numeric", "NULL")) 
```

Convert all variables to factor values and confirm the new class

```{r, echo=TRUE}
AcaciaSaligna_Data$Vegetation_Type <- 
  as.factor(AcaciaSaligna_Data$Vegetation_Type)
class(AcaciaSaligna_Data$Vegetation_Type)

AcaciaSaligna_Data$Year <- as.factor(AcaciaSaligna_Data$Year)
class(AcaciaSaligna_Data$Year)

AcaciaSaligna_Data$Quadrat <- as.factor(AcaciaSaligna_Data$Quadrat)
class(AcaciaSaligna_Data$Quadrat)

AcaciaSaligna_Data$Total_Number <- as.factor(AcaciaSaligna_Data$Total_Number)
class(AcaciaSaligna_Data$Total_Number)

AcaciaSaligna_Data$Total_Density <- as.factor(AcaciaSaligna_Data$Total_Density)
class(AcaciaSaligna_Data$Total_Density)

AcaciaSaligna_Data$Seedling_Number <- as.factor(AcaciaSaligna_Data$Seedling_Number)
class(AcaciaSaligna_Data$Seedling_Number)

AcaciaSaligna_Data$Seedling_Density <- as.factor(AcaciaSaligna_Data$Seedling_Density)
class(AcaciaSaligna_Data$Seedling_Density)

AcaciaSaligna_Data$Adult_Number <- as.factor(AcaciaSaligna_Data$Adult_Number)
class(AcaciaSaligna_Data$Adult_Number)

AcaciaSaligna_Data$Adult_Density <- as.factor(AcaciaSaligna_Data$Adult_Density)
class(AcaciaSaligna_Data$Adult_Density)

summary(AcaciaSaligna_Data)
```



```{r, echo=TRUE}
# Convert factor columns to numeric
AcaciaSaligna_Data <- AcaciaSaligna_Data %>%
  mutate(
    Total_Density = as.numeric(as.character(Total_Density)),
    Seedling_Density = as.numeric(as.character(Seedling_Density)),
    Adult_Density = as.numeric(as.character(Adult_Density))
  )

str(AcaciaSaligna_Data)

# Calculate average densities for each year
summary_data <- AcaciaSaligna_Data %>%
  group_by(Year) %>%
  summarise(
    Total_Density = mean(Total_Density, na.rm = TRUE),
    Seedling_Density = mean(Seedling_Density, na.rm = TRUE),
    Adult_Density = mean(Adult_Density, na.rm = TRUE)
  )


# Reshape data to long format
summary_data_long <- summary_data %>%
  pivot_longer(cols = c(Total_Density, Seedling_Density, Adult_Density),
               names_to = "Group",
               values_to = "Avg_Density")

# Check structure
print(summary_data_long)


# Calculate Averages 
summary_data <- AcaciaSaligna_Data %>%
  group_by(Year, Vegetation_Type) %>%
  summarise(
    Total_Density = mean(Total_Density, na.rm = TRUE),
    Seedling_Density = mean(Seedling_Density, na.rm = TRUE),
    Adult_Density = mean(Adult_Density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(Total_Density, Seedling_Density, Adult_Density),
               names_to = "Group",
               values_to = "Avg_Density")

# Plot bar charts 
ggplot(summary_data, aes(x = factor(Year), y = Avg_Density, fill = Group)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +  # Adjust bar width and spacing
  facet_wrap(~Vegetation_Type, ncol = 1) +  
  labs(x = "Year", y = "Average Species Density", fill = "Density Type") + 
  theme_minimal() +  
  scale_fill_manual(values = c("Total_Density" = "black", 
                               "Seedling_Density" = "grey", 
                               "Adult_Density" = "grey39")) +  
  theme(text = element_text(size = 12), 
        strip.text = element_text(size = 12),  
        axis.text.x = element_text(angle = 45, hjust = 1))  

```

HIER FIGURE CAPTION EINFÜGEN UND ERKLÄREN DASS IN MANCHEN AREAS IN MANCHEN 
JAHREN NICHT GESAMPLED WURDE


