---
title: "Acacia_saligna"
author: "Lilith Diener"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Temporal assessment of the density and population structure of Acacia saligna 
## in the Cape Flats Nature Reserve

Invasive alien plants pose a serious threat to native vegetation. Knowledge of 
changes in the density of invaders is thus crucial to evaluate the effectiveness
of eradication methods. Research on the population density and structure of 
Acacia saligna was conducted between 2019 and 2024 in a previously burnt section 
of the Cape Flats Nature Reserve, and its impact on the local plant species 
richness determined. In seven 4mx4m quadrats each, the number of A. saligna 
individuals was recorded in three areas across the reserve (Ecotone, 
Sand Fynbos, Strandveld). A. saligna plants were cleared and a decline in the 
population's average density was observed.

Let us start with the coding... Firstly, we are installing and loading all
necessary packages using the library() function.

```{r, echo=TRUE}
# Install and load packages
# install.packages("tidyr")
# options("install.lock"=FALSE)
# install.packages("ggplot2")
# options("install.lock"=FALSE)

library(tidyr)
library(dplyr)
library(ggplot2)
```

Now, we can proceed to read in the dataset using read.csv().

```{r, echo=TRUE}
AcaciaSaligna_Data <- read.csv("AcaciaSaligna_Data.csv")
```

I removed the columns "Area" and "Number_other_species" from the dataset as 
they are not needed for this analysis. Note that these two columns are marked
as "NULL".

```{r, echo=TRUE}
AcaciaSaligna_Data <- read.table("AcaciaSaligna_Data.csv", header = TRUE, 
                                 sep = ",", 
		colClasses = c("NULL","integer", "character", "integer", "integer", 
						               "numeric", 
										"integer", "numeric", "integer", "numeric", "NULL")) 
```

Here, I converted all the densities in the dataset to numeric values and 
verified the output using str().

```{r, echo=TRUE}
# Convert factor columns to numeric
AcaciaSaligna_Data <- AcaciaSaligna_Data %>%
  mutate(
    Total_Density = as.numeric(as.character(Total_Density)),
    Seedling_Density = as.numeric(as.character(Seedling_Density)),
    Adult_Density = as.numeric(as.character(Adult_Density))
  )

str(AcaciaSaligna_Data)
```

In the last few steps, I calculated the average total, adult, and seedling
densities for each year and vegetation type. I also put the data into long 
format using pivot_longer().

```{r, echo=TRUE}
# Calculate average densities for each year
summary_data <- AcaciaSaligna_Data %>%
group_by(Year) %>%
summarise(Total_Density = mean(Total_Density, na.rm = TRUE),
Seedling_Density = mean(Seedling_Density, na.rm = TRUE),
Adult_Density = mean(Adult_Density, na.rm = TRUE))

str(AcaciaSaligna_Data) # Check structure of data

# Calculate average densities for each year and vegetation type
summary_data <- AcaciaSaligna_Data %>%
  group_by(Year, Vegetation_Type) %>%
  summarise(
    Total_Density = mean(Total_Density, na.rm = TRUE),
    Seedling_Density = mean(Seedling_Density, na.rm = TRUE),
    Adult_Density = mean(Adult_Density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(Total_Density, Seedling_Density, Adult_Density),
               names_to = "Group",
               values_to = "Avg_Density")
```

Finally, I created three bar charts showing changes in the average total, 
seedling and adult density of A. saligna per square meter in each area from 2019 
to 2024. Please note that the Ecotone and Sand Fynbos areas were only sampled in 
some of the years. 

```{r, echo=TRUE}
# Plot bar charts 
ggplot(summary_data, aes(x = factor(Year), y = Avg_Density, fill = Group)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) + 
  facet_wrap(~Vegetation_Type, ncol = 1) +  
  labs(x = "Year", y = "Average Species Density", fill = "Density Type") + 
  theme_minimal() +  
  scale_fill_manual(values = c("Total_Density" = "black", 
                               "Seedling_Density" = "grey", 
                               "Adult_Density" = "grey39")) +  
  theme(text = element_text(size = 12), 
        strip.text = element_text(size = 12),  
        axis.text.x = element_text(angle = 45, hjust = 1))  
```




